---
title: "Multiple linear regression"
output: 
    bookdown::html_document2
css: "../CSS/lab.css"
date: 'Last compiled: `r format(Sys.time(), "%b %d, %Y")`'
bibliography: ["ProfBeauty.bib", "PackagesUsed.bib"]
---

<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). 
This lab was written by Mine &Ccedil;etinkaya-Rundel and Andrew Bray and modified by Alan Arnholt.
</div>

```{r label = "setup", include = FALSE}
knitr::opts_chunk$set(comment = NA, fig.align = "center", warning = FALSE, message = FALSE)
```

<div id="instructions">
Complete all **Exercises and Questions**.  Submit your knitted html file to [crowdgrader](http://crowdgrader.com).
</div>

## Grading the professor {-}

Many college courses conclude by giving students the opportunity to evaluate 
the course and the instructor anonymously. However, the use of these student 
evaluations as an indicator of course quality and teaching effectiveness is 
often criticized because these measures may reflect the influence of 
non-teaching related characteristics, such as the physical appearance of the 
instructor. The article titled, "Beauty in the classroom: instructors' 
pulchritude and putative pedagogical productivity" [@Hamermesh2005] 
found that instructors who are judged more attractive receive higher 
instructional ratings. 

In this lab, we will analyze the data from this study in order to learn what goes 
into a positive professor evaluation.

## Getting Started {-}

### Load packages {-}

```{r, echo = FALSE, results = 'hide'}
PackagesUsed <- c("DT", "ggplot2", "knitr", "rmarkdown", "dplyr", "GGally", "statsr", "htmlTable")
# Write bib information
knitr::write_bib(PackagesUsed, file = "./PackagesUsed.bib")
# Load packages
lapply(PackagesUsed, library, character.only = TRUE)
```

 

In this lab, the `dplyr` package by @R-dplyr is used to explore the data.  Visualization of the data 
is accomplished with the `ggplot2` and `GGally` packages created by @R-ggplot2 and @R-GGally, respectively. The data resides in the `statsr` package created by @R-statsr.  Tables are formatted for html with the `htmlTable` package created by @R-htmlTable and data is formatted for easy viewing with the `DT` package by @R-DT.

Let us load the packages.

```{r label = "load-packages"}
library(statsr)
library(dplyr)
library(ggplot2)
library(GGally)
library(htmlTable)
library(DT)
```

This is the first time we have used the `GGally` package. We will use the
`ggpairs` function from `GGally` later in the lab.

### The data {-}

The data were gathered from end of semester student evaluations for a large 
sample of professors from the University of Texas at Austin. In addition, six 
students rated the professors' physical appearance. (This is a slightly modified 
version of the original data set that was released as part of the replication 
data for *Data Analysis Using Regression and Multilevel/Hierarchical Models* 
[@Gelman2007].) The result is a data frame where each row contains a 
different course and columns represent variables about the courses and 
professors.

Let us load the data:

```{r label = "load-data"}
data(evals)
```

```{r, label = "TableDesc", echo = FALSE}
# options(table_counter = TRUE)
TRY <- data.frame(variable = c("score", "rank", "ethnicity", "gender", "language", "age",
                               "cls_perc_eval", "cls_did_eval", "cls_students", "cls_level",
                               "cls_profs", "cls_credits", "bty_f1lower", "bty_f1upper",
                               "bty_m1lower", "bty_m1upper", "bty_m2upper",
                               "bty_avg", "pic_outfit", "pic_color"),
                  description = c("average professor evaluation score: (1) very unsatisfactory---(5) excellent.", "rank of professor: teaching, tenure track, tenured.",
                                  "ethnicity of professor: not minority, minority.",
                                  "gender of professor: female, male.",
                                  "language of school where professor received education: english or non-english.",
                                  "age of professor.",
                                  "percent of students in class who completed evaluation.",
                                  "number of students in class who completed evaluation.",
                                  "total number of students in class.",
                                  "class level: lower, upper.",
                                  "number of professors teaching sections in course in sample: single, multiple.",
                                  "number of credits of class: one credit (lab, PE, etc.), multi credit.",
                                  "beauty rating of professor from lower level female: (1) lowest---(10) highest.",
                                  "beauty rating of professor from upper level female: (1) lowest---(10) highest.",
                                  " beauty rating of professor from lower level male: (1) lowest---(10) highest.",
                                  "beauty rating of professor from upper level male: (1) lowest---(10) highest.",
                                  "beauty rating of professor from second upper level male: (1) lowest---(10) highest.",
                                  "average beauty rating of professor.",
                                  "outfit of professor in picture: not formal, formal.",
                                  "color of professor's picture: color, black & white."))
# htmlTable::htmlTable(as.matrix(TRY), align = "ll", css.cell = "padding-left: 1em; padding-right: 1em;", caption = "Variables and descriptions for `evals`")

knitr::kable(as.matrix(TRY), align = "ll", caption = "Variables and descriptions for `evals`")
```


```{r, label = "show"}
DT::datatable(evals)
```

1. Is this an observational study or an experiment?
<ol>
<li> Observational study </li>
<li> Experiment </li>
</ol>

<div id="answer">
Type your answer for Question 1 between the answer tags and delete this comment.


</div>

2. The original research question posed in the paper is whether beauty leads directly to the differences in course evaluations.  Given the study design, should the question be rephrased? If so, how? 
<ol>
<li> No, the question is worded accurately. </li>
<li> Yes, revise wording to "Is there an association between beauty and course evaluations?" </li>
<li> Yes, revise wording to "Does beauty score increase the professor's course evaluations?"  </li>
<li> Yes, revise wording to "Does beauty score decrease the professor's course evaluations?" </li>
</ol>

<div id="answer">
Type your answer for Question 2 between the answer tags and delete this comment.


</div>

## Exploring the data {-}

3. Which of the following statements is **false** about the distribution of `score`? 
<ol>
<li> The median of the distribution is 4.3. </li>
<li> 25\% of the students gave their professors a score of over 4.6. </li> 
<li> 11 of students gave a professor a score below 3. </li> 
<li> The left skewness of the data suggests that the students are less likely to rate the professors highly. </li>
</ol>

```{r label = "score-dist"}
# Type your code for Question 3 here, and Knit.


```

<div id="answer">
Type your answer for Question 3 between the answer tags and delete this comment.


</div>

<div id="exercise">
**Exercise 1:** Excluding `score`, select two other variables and describe their 
relationship using an appropriate visualization (scatterplot, side-by-side boxplots, 
or mosaic plot).
</div>

```{r label = "two-vars-eda", fig.width = 5, fig.height = 5}
# Type your code for the Exercise 1 here, and Knit.

```

<div id="answer">
Type your answer for Exercise 1 between the answer tags and delete this comment. - Answers will vary


</div>

## Simple linear regression {-}

The fundamental phenomenon suggested by the study is that teachers judged more attractive
are evaluated more favorably. Let us create a scatterplot to see if this appears 
to be the case:

```{r label = "scatterSBA", fig.cap = "Score versus beauty average"}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_point(color = "lightblue") + 
  theme_bw() + 
  labs(x = "average beauty score", y = "evaluation score")
```

Before we draw conclusions about the trend in Figure \@ref(fig:scatterSBA), compare the number of observations 
in the data frame with the approximate number of points on the scatterplot. 
Is anything awry?

<div id="answer">
Type your answer here - answers may vary


</div>

<div id="exercise">
**Exercise 2:** Replot the scatterplot, but this time replace the `geom_point()` layer 
with a `geom_jitter()` layer. (Use `?geom_jitter` to learn more.)  What was misleading 
about the initial scatterplot?
</div>

```{r label = "jitter"}
# Type your code for the Exercise 2 here, and Knit.

```

<div id="answer">
Type your answer for Exercise 2 between the answer tags and delete this comment. - Answers will vary


</div>

Is the apparent trend in the plot something more than 
natural variation? Fit a linear model called `modB` to predict average professor 
score by average beauty rating and add the line to your plot using the following. If 
you do not remember how to do this, refer to the previous lab.  

```{r label = "scatter-score-bty_avg-line"}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter(color = "lightblue", alpha = 0.8) +
  geom_smooth(method = "lm") + 
  theme_bw() +
  labs(x = "average beauty score", y = "evaluation score")
```

The blue line is the model. The shaded gray area around the line tells us about the 
variability we might expect in our predictions. To remove the confidence band, use `se = FALSE`.

```{r label = "scatter-score-bty_avg-line-no-se"}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_jitter(color = "lightblue", alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) + 
  theme_bw() + 
  labs(x = "average beauty score", y = "evaluation score")
```



<div id="exercise">
**Exercise 3:** Print a summary of the linear model, write out the equation, and 
interpret the slope. 
</div>

```{r label = "summary-model"}
# Type your code for the Exercise 3 here, and Knit.


```

<div id="answer">
Type your answer for Exercise 3 between the answer tags and delete this comment.



</div>

4. Average beauty score is a statistically significant predictor of evaluation score. 
<ol> 
<li> True </li>
<li> False </li>
</ol>

<div id="answer">
Type your answer for Question 4 between the answer tags and delete this comment.


</div>

5. Use residual plots to evaluate whether the conditions of least squares regression are reasonable.  Which of the following statements is an incorrect analysis of the residual plots and conditions? 
<ol>
<li> Linear association: The residuals plot shows a random scatter. </li>
<li> Constant variance of residuals: No fan shape in residuals plot. </li>
<li> Nearly normal residuals: Residuals are right skewed, but the sample size is large, so this may not be an important violation of conditions. </li>
<li> Independent observations: Classes sampled randomly, no order effect. </li>
</ol>

```{r label = "diag-score-bty_avg", fig.width = 13.5}
# Type your code for the Question 5 here, and Knit.
# residual plot, Q-Q plot, and density plot of residuals

```

<div id="answer">
Type your answer for Question 5 between the answer tags and delete this comment.


</div>

## Multiple linear regression {-}

The data set contains several variables on the beauty score of the professor as well as 
individual ratings from each of the six students who were asked to score the 
physical appearance of the professors. It also has a variable that is the average of these six scores. Let us 
examine the strength of the linear relationship between one of these scores 
(`bty_f1lower`) and the average beauty score (`bty_avg`).

```{r label = "bty-rel"}
ggplot(data = evals, aes(x = bty_f1lower, y = bty_avg)) +
  geom_jitter()
# Correlation using cor()
# 1. Just the function
cor(evals$bty_avg, evals$bty_f1lower)
# 2. Using with() to access the data
with(data = evals, cor(bty_avg, bty_f1lower))
# 3. using a pipe
evals %>% 
  summarise(r = cor(bty_avg, bty_f1lower))
```

As expected, the relationship is quite strong---after all, the average score is 
calculated using the individual scores. Figure \@ref(fig:bty-rels) depicts the 
relationships between all beauty variables (columns 13 through 19) using the 
following command:

```{r label = "bty-rels", fig.width = 12, fig.height= 8, fig.cap = "GGpairs matrix scatterplot"}
# Note: This is a painfully slow function!
ggpairs(evals, columns = 13:19)
```

The variables in Figure \@ref(fig:bty-rels) are collinear (correlated), and adding more than one of these 
variables to the model would not add much value to the model. In this 
application and with these highly-correlated predictors, it is reasonable to use
the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score 
after we have accounted for the gender of the professor, we can add the gender 
term into the model.

```{r label = "scatter-score-bty_avg_gender"}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```

```{r, label = "summMBTYGEN", echo = FALSE}
options(table_counter = 1)
htmlTable::htmlTable(round(summary(m_bty_gen)$coef, 4), css.cell = "padding-left: 1em; padding-right: 1em;", caption ="`score` regressed on `bty_avg` and `gender`", align = "rrrr")
```

6. P-values and parameter estimates should only be trusted if the conditions for the regression are reasonable. Using diagnostic plots, we can conclude that the conditions for this model (`m_bty_gen`) may be reasonable. 
<ol>
<li> True </li>
<li> False </li>
</ol>

```{r label = "diag-mlr-model", fig.width = 13.5}
# Type your code to create appropriate diagnostic plots, and Knit.


```

<div id="answer">
Type your answer for Question 6 between the answer tags and delete this comment.


</div>

<div id="exercise">
**Exercise 4:** Print a summary of the multiple linear regression model (`m_bty_gen`).  Is `bty_avg` 
still a significant predictor of `score`?  Has the addition of `gender` to the 
model changed the parameter estimate for `bty_avg`?
</div>

```{r label = "summary-mlr-model"}
# Type your code for Exercise 4 here, and Knit.


```

<div id="answer">
Type your answer for Exercise 4 between the answer tags and delete this comment.


</div>


Note that the estimate for `gender` is now called `gendermale`. The original variable 
name will change whenever you introduce a categorical variable. The reason the name 
changes is because R re-codes `gender` from having values `female` and `male` to an 
indicator variable called `gendermale` that takes a value of $0$ for females and
a value of $1$ for males. (Such variables are often referred to as "dummy" 
variables.)

As a result, for females, the parameter estimate is multiplied by zero, leaving 
the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{\text{score}} &= \hat{\beta}_0 + \hat{\beta}_1 \times \text{bty_avg} + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times \text{bty_avg}\end{aligned}
\]

7. For two professors (one male and one female) who received the same beauty rating, the male professor is predicted to have the higher course evaluation score than the female.
<ol>
<li> True </li>
<li> False </li>
</ol>

<div id="answer">
Type your answer for Question 7 between the answer tags and delete this comment.


</div>

The decision to call the indicator variable `gendermale` instead of`genderfemale`
has no deeper meaning. R simply codes the category that comes first 
alphabetically as a $0$. (You can change the reference level of a categorical 
variable, which is the level that is coded as a 0, using the`relevel` function; use `?relevel` to learn more.)

<div id="exercise">
**Exercise 5:** Create a new model called `m_bty_rank` with `gender` removed and 
`rank` added to the model.  How does R handle categorical variables that have 
more than two levels? Note that the rank variable has three levels: teaching, 
tenure track, and tenured.
</div>

```{r label = "new-mlr-model"}
# Type your code for Exercise 5 here, and Knit.


```

<div id="answer">
Type your answer for Exercise 5 between the answer tags and delete this comment.


</div>

8. Which of the following is the correct order of the three levels of rank if we were to order them from lowest predicted course evaluation score to highest predicted course evaluation score? 
<ol>
<li> Teaching, Tenure Track, Tenured </li>
<li> Tenure track, Tenured </li>
<li> Tenure Track, Tenured, Teaching </li>
<li> Teaching, Tenured, Tenure Track </li>
</ol>

<div id="answer">
Type your answer for Question 8 between the answer tags and delete this comment.



</div>

The interpretation of the coefficients in multiple regression is slightly 
different from that of simple regression. The estimate for `bty_avg` reflects
how much higher a group of professors is expected to score if they have a beauty
rating that is one point higher *while holding all other variables constant*. In
this case, that translates into considering only professors of the same rank 
with `bty_avg` scores that are one point apart.

## Prediction {-}

Suppose we want to use the model we created earlier, `m_bty_gen`, to predict 
the evaluation score for a professor, Dr. Hypo Thetical, who is a male tenure track
professor with an average beauty score of 3.

If we wanted to do this by hand, we would "plug in" these values using the 
fitted linear model.

To compute the predicted value in R, create a new data frame for this professor.

```{r label = "new-prof"}
newprof <- data.frame(gender = "male", bty_avg = 3)
```

Note that I did not need to add `rank = "tenure track"` to this data frame since
this variable is not used in our model.

Then, I can do the prediction using the `predict` function:

```{r label = "new-prof-predict"}
predict(m_bty_gen, newdata = newprof)
```

We can also construct a prediction interval around this prediction, which will
provide a measure of uncertainty around the prediction.

```{r label = "new-prof-predict-interval"}
PI <- predict(m_bty_gen, newdata = newprof, interval = "prediction", level = 0.95)
PI
```

The model predicts, with 95% confidence, that a male professor with an 
average beauty score of 3 is expected to have an evaluation score between `r PI[2]` 
and `r PI[3]`. 

## The search for the best model {-}

We will start with a full model that predicts professor score based on rank, 
ethnicity, gender, language of the university where they got their degree, age, 
proportion of students that filled out evaluations, class size, course level, 
number of professors, number of credits, average beauty rating, outfit, and 
picture color.

<div id="exercise">
**Think About**: Which variable would you expect to have the highest p-value in this model? Why? 
*Hint:* Think about which variable would you expect to not have any association 
with the professor score.
</div>


Let us run the model...

```{r m_full, tidy = FALSE}
m_full <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

```{r, label = "FULLMOD", echo = FALSE}
htmlTable::htmlTable(round(summary(m_full)$coef, 4), css.cell = "padding-left: 1em; padding-right: 1em;", caption ="Full Model", align = "rrrr")
```


9. Which of the following is the correct interpretation of the coefficient associated with the ethnicity variable. <br> Non-minority professors are expected on average to score ... 
<ol>
<li> 0.12 points lower than minority professors, all else held constant. </li>
<li> 0.12 points higher than minority professors, all else held constant. </li> 
<li> 0.08 points lower than minority professors, all else held constant. </li>
<li> 0.08 points higher than minority professors, all else held constant. </li>
</ol>

<div id="answer">
Type your answer for Question 9 between the answer tags and delete this comment.


</div>

<div id="exercise">
**Exercise 7:** Drop the variable with the highest p-value and re-fit the model. 
Did the coefficients and significance of the other explanatory variables change? 
(One of the things that makes multiple regression interesting is that coefficient 
estimates depend on the other variables that are included in the model.) If not, 
what does this say about whether or not the dropped variable was collinear with 
the other explanatory variables?
</div>

```{r label = "p-val-select"}
# Type your code for Exercise 7 here, and Knit.


```

<div id="answer">
Type your answer for Exercise 7 between the answer tag and delete this comment.


</div>

Next we use a different model selection technique: adjusted $R^2$. Create a new model, 
`m1`, where you remove `rank` from the list of explanatory variables. Compare the 
adjusted $R^2$ of this new model to the adjusted $R^2$ of the full model.

```{r eval=FALSE}
m1 <- lm(score ~ ethnicity + gender + language + age + cls_perc_eval + 
           cls_students + cls_level + cls_profs + cls_credits + 
           bty_avg, data = evals)
summary(m1)$adj.r.squared
```

Then, try dropping the next variable from the full model (`ethnicity`):

```{r eval=FALSE}
m2 = lm(score ~ rank + gender + language + age + cls_perc_eval + cls_students + 
          cls_level + cls_profs + cls_credits + bty_avg, data = evals)
summary(m2)$adj.r.squared
```

<div id="exercise">
**Exercise 8:** Repeat this process until you have removed each variable, one at a time, 
from the full model. Which variable yields the highest improvement in the adjusted $R^2$?
</div>

```{r label = "adj-rsq-select"}
# Type your code for Exercise 8 here, and Knit.


```

<div id="answer">
Type your answer for Exercise 8 between the answer tags and delete this comment.


</div>


10. Elimination of which variable from the full model yielded the highest adjusted R-squared? 
<ol>
<li> `bty_avg` </li>
<li> `cls_profs` </li> 
<li> `cls_students` </li>
<li> `rank` </li>
</ol>

<div id="answer">
Type your answer for Question 10 between the answer tags and delete this comment.


</div>

To complete the model selection, we would continue removing variables one at a 
time until removal of another variable did not increase adjusted $R^2$.

```{r label = "final_model"}
# Type your code for the final model, and Knit.

```


<div id="exercise">
**Exercise 9:** The original paper describes how these data were gathered by taking 
a sample of professors from the University of Texas at Austin and including all 
courses that they have taught.  Considering that each row represents a course, 
could this new information have an impact on any of the conditions of linear regression?
</div>

<div id="answer">
Type your answer for Exercise 9 between the answer tags and delete this comment.


</div>

<div id="exercise">
**Exercise 10:** Based on your final model, describe the characteristics of a 
professor and course at University of Texas at Austin that would be associated 
with a high evaluation score.
</div>

<div id="answer">
Type your answer for Exercise 10 between the answer tags and delete this comment.



</div>

<div id="exercise">
**Exercise 11:** Would you be comfortable generalizing your conclusions to apply 
to professors generally (at any university)?  Why or why not?
</div>

<div id="answer">
Type your answer for Exercise 11 between the answer tags and delete this comment.


</div>

```{r, label = "sessioninfo"}
sessionInfo()
```

## References {-}